---
title: "Lab 5"
author: "Sophie McKee"
date: "`r Sys.Date()`"
output: pdf_document
---

#========================================================
# A. Set up
#========================================================

1.Set up and clean global environment. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clean existing variables
rm(list=ls())
```


2.Load required libraries (to read excel files, to manipulate dataframes, to plot time series)

```{r}

library(readxl)
library(dplyr)
library(tseries)
library(tidyverse)
library(Metrics)

```

#========================================================
# B. Define model variables
#========================================================

```{r}
# Input the latest version of the file "FinalData2026.xlsx

data <- read_excel("ProjectData/FinalData2026-2.xlsx") %>% as.data.frame()


# Create a standardized SLead_SalesTax variable

data_1 <- data %>% mutate(SLead_SalesTax = ifelse(Year < 2024, 
                                                  Lead_SalesTax * (4.35 / 3.85),
                                                  Lead_SalesTax)) %>% 
  filter(Year >= 2015, Year <= 2025) %>% dplyr::select(MAN, RT, LH, EDUHS, SLead_SalesTax, CPI) %>% na.omit()


data_1
# Import your external data

unem <- read_excel("ProjectData/FORT608URN.xlsx", sheet = "Monthly") %>% 
  as_tibble() %>% dplyr::select(FORT608URN)


# Create the dataframe df (make sure to remove any row containing NAs)

df <- bind_cols(data_1, unem)

df$FORT608URN[130] <- 3.4


# Create a data frame ldf by logging the dataframe df

ldf <- ts(log(df))

```




#========================================================
# C. Create VECM model using data up to 2025/12
#========================================================

Calling the packages for use

```{r}
library(urca)  # ca.jo
library(tsDyn) # VECM
library(vars)  # vec2var
```

Johansen procedure 

```{r}
Johansen <- ca.jo(ldf, ecdet = "none", type = "eigen", K = 12)
summary(Johansen)
```
Confirm the number of cointegrating vectors based on the results of the Johansen procedure.

NOTE: You need at least 4 cointegrating vectors.

Print the matrix of cointegrating vectors.

```{r}
nvectors = 4
Beta <- round(coefB(Johansen, r = nvectors), 3)
Beta
```


Short run adjustment coefficients

```{r}
vecm <- VECM(ldf, lag = 11, r = nvectors, estim = "ML", include = "const")
summary(vecm)

```

For each vector, we need at least one adjustment speed to have the oppositive sign of the corresponding cointegrating vector coefficient and be statistically significant.

Vec 1: MAN has loading -0.7203*** (opposite sign to cointegrating coeff +1.0). Error correction confirmed.
Vec 2: FORT608URN has loading -12.8842* (opposite sign to cointegrating coeff +0.206). Error correction confirmed.
Vec 3: CPI has loading -0.0247* (opposite sign to cointegrating coeff +8.663). Error correction confirmed.
Vec 4: FORT608URN loading is -0.7390 (correct opposite sign vs +0.323) but not significant. This vector lacks strong error correction support.

....

#========================================================
# D. Test if model predicts 2025 data well
#========================================================

Rerun the same model on data up to 2024/12 (included)

```{r}
# truncate the data by removing the 12 last observations
numrow <- nrow(ldf)
ldf2024 <- ldf[1:(numrow- 12),]
# Fit the same VECM model
Johansen2024 <- ca.jo(ldf2024, ecdet="none", type="eigen", K=12)

```

Forecast 2025 data

```{r}
# vec2var: Transform a VECM to VAR in levels
vec2var_ca.jo2024 <- vec2var(Johansen2024, r = nvectors)
# forecasting horizon
nhor <- 12 
# Forecasting
pred_vec2var_ca.jo2025 <- predict(vec2var_ca.jo2024, n.ahead=nhor)
```

Compare actual and forecasted 2025 log of SLead_SalesTax
Calculate Mean Absolute Error for the 12 observations of 2025

```{r}
Forecast2025 <- pred_vec2var_ca.jo2025$fcst$SLead_SalesTax[,1]
Actual2025 <- ldf[(nrow(ldf) - 11): nrow(ldf), "SLead_SalesTax"]

AE <- 
MAE <- mae(Actual2025, Forecast2025)
MAE


```

You can redo steps A-D many times until you find the model with the lowest MAE. 
Play with the variables included in your model and with the number of months included in your dataset. You can try including data starting in 2013, or in 2014, or later, instead of 2012, and see if you get better predictions.

#========================================================
# E. Forecast lead_STF_Real for the months 1/26 to 11/26 and save it.
#========================================================

```{r}
# vec2var: Transform a VECM to VAR in levels
# This is the original model, using all the data
vec2var_ca.jo <- vec2var(Johansen, r = nvectors)
# forecasting horizon
nhor <- 11
# Forecasting 2026 (11 observations)
pred_vec2var_ca.jo <- predict(vec2var_ca.jo, n.ahead=nhor)
Forecast2026 <- as.data.frame(pred_vec2var_ca.jo$fcst$FinalData.lead_STF_Real[,1])
write_csv(Forecast2026, "Forecast2026.csv")



pred_vec2var_ca.jo$fcst$SLead_SalesTax[,1]
```
