---
title: "Lab 4"
author: "Vincent Ma"
date: "`r Sys.Date()`"
output: pdf_document
---

Hint: Use the code provided in "VECM Model example.Rmd"

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# clean existing variables
rm(list=ls())
```

Load required libraries (to read excel files, to manipulate dataframes, to plot time series, to test for stationarity)

```{r}

library(readxl)
library(dplyr)
library(tseries)

```

#========================================================
# Data
#========================================================

Input project data.
Create a variable SLead_SalesTax standardizing Lead_SalesTax as if the sales tax rate had been the same as in 2024 for all the years.[Note: sales tax rate was 3.85% before 2024 and 4.35% starting in 2024]

```{r}
data <- (read_excel("ProjectData/FinalData2026-1.xlsx") %>% as.data.frame()) %>% mutate(SLead_SalesTax = ifelse(Year < 2024, Lead_SalesTax * (4.35 / 3.85), Lead_SalesTax))
```

Plot the log of each of the BLS time series and compare them with the log of SLead_SalesTax to see if they follow similar trends/seasonality.

```{r}

mlc <- ts(log(data$MLC)) ; man <- ts(log(data$MAN)) ; wt <- ts(log(data$WT)) 

rt <- ts(log(data$RT)) ; twu <- ts(log(data$TWU)) ; info <- ts(log(data$INFO)) 

fa <- ts(log(data$FA)) ; pbs <- ts(log(data$PBS)) ; eduhs <- ts(log(data$EDUHS)) 

lh <- ts(log(data$LH)) ; os <- ts(log(data$OS)) ; g <- ts(log(data$G)) 

slead <- ts(log(data$SLead_SalesTax)) 


all <- tibble(slead = slead, mlc = mlc, man = man , wt = wt, rt = rt, twu = twu,  

              info = info, fa = fa, pbs = pbs, eduhs = eduhs) 

all_2 <- tibble(lh = lh, os = os, 

              g = g) 

 

 

plot.ts(all) ; plot.ts(all_2) 

```

Using the information collected in Part I of the lab and the insight you gained from the plotting exercise, create a data frame df including 
- the supply-side BLS series you selected (you can also create aggregates of individual series)
- the demand-side BLS series you selected (you can also aggregate (add up) individual series)
- license counts (if you wish)
- CPI (if you wish)
- SLead_SalesTax

To focus on recent trends, consider filtering the dataset to remove the earliest years of the series.

Create a data frame ldf by logging df, ensuring it doesn't contain any empty observations.

```{r}

unem <- read_excel("ProjectData/FORT608URN.xlsx", sheet = "Monthly") %>% 
  as.tibble() %>% na.omit() %>% 
  select(FORT608URN)

unem_ts <- ts(log(unem))

adf.test(unem_ts)
adf.test(diff(unem_ts))

df <- data %>%
  filter(Year >= 2015) %>%
  select(MAN, PBS, RT, LH, EDUHS, SLead_SalesTax) %>%
  na.omit()

ldf <- ts(log(df))

df_unem <- bind_cols(df, unem = unem) %>%
  rename(unem = FORT608URN)

ldf <- ts(log(df_unem))
```

# Plot the time series included in your model.

```{r}
plot.ts(ldf)
```

# Check the order of integration of each of the time series included in the model.

```{r}
adf.test(ldf[, "MAN"])
adf.test(ldf[, "PBS"])
adf.test(ldf[, "RT"])
adf.test(ldf[, "LH"])
adf.test(ldf[, "EDUHS"])
adf.test(ldf[, "SLead_SalesTax"])
adf.test(ldf[, "unem"])
```

```{r}
# ADF test on first differences
adf.test(diff(ldf[, "MAN"]))
adf.test(diff(ldf[, "PBS"]))
adf.test(diff(ldf[, "RT"]))
adf.test(diff(ldf[, "LH"]))
adf.test(diff(ldf[, "EDUHS"]))
adf.test(diff(ldf[, "SLead_SalesTax"]))
adf.test(diff(ldf[, "unem"]))
```
All series are I(1): non-stationary in levels, stationary in first differences.

#========================================================
# Cointegrating vectors (long run)
#========================================================

Hint: check if your model is acceptable by using the document "Cookbook.docx"

Calling the packages for use

```{r}
library(urca)
library(tsDyn)
```

# Conduct the Johansen procedure

```{r}
Johansen <- ca.jo(ldf, ecdet = "none", type = "eigen", K = 12)
summary(Johansen)
```

Number of cointegrating vectors: 4 (at 5%)

Coefficients of the cointegrating vectors

```{r}
Beta <- round(coefB(Johansen, r = 4), 3)
Beta
```

#========================================================
# Short run adjustment coefficients
#========================================================

```{r}
vecm <- VECM(ldf, lag = 11, r = 4, estim = "ML", include = "const")
summary(vecm)

```

For each vector, we need at least one adjustment speed to have the oppositive sign of the corresponding cointegrating vector coefficient and be statistically significant.[Hint: check the R code VECM Model example.Rmd and the document cookbook.docx]

Vec 1: For the first vector, MAN clears the vector because the coefficient on MAN is negative (-0.7324) and the cointegrating vector is positive (1), and MAN is significant (***).

Vec 2: For the second vector, PBS clears the vector because the coefficient on PBS is negative (-0.3122) and the cointegrating vector is positive (1), and PBS is significant (*).

Vec 3: For the third vector, EDUHS clears the vector because the coefficient on EDUHS is positive (1.4265) and the cointegrating vector is negative (-0.713), and EDUHS is significant (*).

Vec 4: For the fourth vector, LH clears the vector because the coefficient on LH is negative (-1.1163) and the cointegrating vector is positive (1), and LH is significant (**).
....

